Rectangulotopes

### 2024-04-23 draft/rectangulotopes.tex
minor changes

% Loday and anti-Loday accents (?)
\makeatletter
\newcommand{\oset}[3][0ex]{%
  \mathrel{\mathop{#3}\limits^{
    \vbox to#1{\kern-2\ex@
    \hbox{$\scriptstyle#2$}\vss}}}}
%
\newcommand{\uset}[3][0ex]{%
  \mathrel{\mathop{#3}\limits_{
    \vbox to#1{\kern-7\ex@
    \hbox{$\scriptstyle#2$}\vss}}}}
\makeatother
%
\newcommand{\loday}[1]{\smash{\overset{\frown}{#1}}}
\newcommand{\antiloday}[1]{\smash{\overset{\smile}{#1}}}
\newcommand{\upArc}[1]{\smash{\raisebox{.05cm}{$\uset[0ex]{#1}{\frown}$}}}
\newcommand{\downArc}[1]{\smash{\raisebox{-.05cm}{$\oset[.2ex]{#1}{\smile}$}}}
% yin and yang accents (?)
\newcommand{\yin}[1]{\smash{\overset{\sim}{#1}}}
\newcommand{\yang}[1]{\smash{\overset{\backsim}{#1}}}
\newcommand{\yinArc}[2]{\smash{\raisebox{.05cm}{$\uset[0ex]{#1}{\frown}$}\hspace{-.3ex}\raisebox{-.03cm}{$\oset[.2ex]{#2}{\smile}$}}}
\newcommand{\yangArc}[2]{\smash{\raisebox{-.03cm}{$\oset[.2ex]{#1}{\smile}$}\hspace{-.3ex}\raisebox{.05cm}{$\uset[0ex]{#2}{\frown}$}}}
% weak and strong equivalence on S_n
\newcommand{\weakeq}{\asymp}
\newcommand{\strongeq}{\mathbin{\smash{\begin{smallmatrix} \backsim \\[-.3cm] \sim \end{smallmatrix}}}}% % J: I would like to stack \backsim and \sim instead

@@

Given a rectangulation~$R$, let us consider the source and target trees~$S(R)$ and~$T(R)$, and orient the horizontal edges from left to right, and the vertical edges from bottom to top.
This defines two partial orders on $[n]$.
The \defn{weak poset}~$P_w(R)=([n],\prec_w)$ of the rectangulation~$R$ is the transitive closure of the union of these two partial orders defined by~$S(R)$ and~$T(R)$.
Equivalently, the weak poset can be defined by considering the adjacency graph of the rectangles in a diagonal representative of the weak equivalence class of $R$, orienting its edges from left to right and from bottom to top, and taking the transitive closure of this directed acyclic graph.
These posets were characterized (and called Baxter posets) by E.~Meehan~\cite{MR4014603}.
The weak poset is a two-dimensional lattice, whose minimum is the root of~$S(R)$ and whose maximum is the root of~$T(R)$.
See \cref{fig:weakPoset}.

@@

We call this congruence the \defn{weak rectangulation congruence} (it is sometimes referred to as the \defn{Baxter congruence}).
The \defn{weak rectangulotope} $\WRP(n)$ is the quotientope of the weak rectangulation congruence $
\weakeq$ on the weak Bruhat order on $\f{S}_n$.

\vincent{Add here sentence about geometric interpretation of weak poset. Bijection cover relations weak poset -- facets -- flips...}

@@

These arcs can be classified into two families of arcs: the up arcs and the down arcs.
For an interval $I$ of $[n]$ of size at least~$2$, the \defn{up arc}~$\upArc{I}$ (resp.~\defn{down arc}~$\downArc{I}$) is the arc that starts at~$\min(I)$ and ends at~$\max(I)$, and passes above (resp.~below) all remaining elements of~$I$, that is
\begin{align*}
\upArc{I} & \eqdef \big( \min(I), \max(I), I \ssm \{\min(I), \max(I)\}, \varnothing \big) \\
\qquad\text{and}\qquad
\downArc{I} & \eqdef \big( \min(I), \max(I), \varnothing, I \ssm \{\min(I), \max(I)\} \big).
\end{align*}
Note that the basic arc~$(i, i+1, \varnothing, \varnothing)$ is both an up and down arc.

@@

Given a rectangulation $R$, we define the \defn{strong poset} $P_s(R)=([n],\prec_s)$ of $R$ by considering two relations on $[n]$~\cite{ACFF24}:
\begin{enumerate}
\item The \defn{adjacency poset} $([n],\tri)$ of $R$, defined by considering the adjacency graph of the rectangles of $R$ (identified by their inorder labels in $[n]$), orienting its edges from left to right and from bottom to top, and taking the transitive closure of this directed acyclic graph.
\item The \defn{special relation} $([n],\btri)$ of $R$, in which $a\btri b$ if the two rectangles $r_a$ and $r_b$ with respective inorder labels $a$ and $b$ are such that 
	\begin{enumerate}[(i)]
	\item either the right side of $r_b$ and the left side of $r_a$ lie on the same vertical segment, and the bottom-right corner of $r_b$ is above the top-left corner of $r_a$, 
	\item or the top side of $r_b$ and the bottom side of rectangle $r_a$ lie on the same horizontal segment, and the top-left corner of $r_b$ lies on the right of the bottom-right corner~of~$r_a$.
	\end{enumerate}
\end{enumerate}
The strong poset is the transitive closure of the union of the two relations $\tri$ and $\btri$.
See \cref{fig:strongPoset}.
The strong poset is a two-dimensional lattice, whose minimum is the root of~$S(R)$ and whose maximum is the root of~$T(R)$. The following statement will be used later.

@@

@@

\cref{prop:strongMinkowski} gives us an alternative description of $\SRP(n)$ as the sum of two opposite quotientopes defined by the \defn{yin} and \defn{yang congruences} respectively, where the yin (resp.~yang) congruence is defined by the arc ideal consisting of all yin (resp.~yang) arcs.
It can be checked that the equivalence classes of these congruences are counted by \defn{semi-Baxter permutations}, defined as permutations avoiding $2\underline{41}3$, hence only one of the two forbidden Baxter patterns.
These equivalence classes are in turn in bijection with rectangulations that avoid one of the two patterns defining diagonal rectangulations, see~\cite{ACFF24}.

@@

\begin{lemma}
  \label{lem:yincount}
  Let $R$ be a strong rectangulation of size~$n$, let~$S$ and $T$ be its source and target trees, let~$([n],\prec_s)$ be its strong poset, and let~$\b{p}(R)$ be the vertex of $\SRP(n)$ corresponding to $R$.
  Given a pair $i,j\in [n]$ with $i<j$, the number of yin arcs $\alpha$ such that $\b{e}_i-\b{e}_j$ is the extremal vertex of~$\SP(\alpha)$ contributing to $\b{p}(R)$ is
  \[
    h^T_i \cdot cv^{T,S}_{i,j}\cdot h^S_j 
  \]
  if $i\succ_s j$, and $0$ otherwise, where
  \begin{itemize}
  \item $h^T_i$ denotes the number of leaves in the horizontal and vertical subtrees of $i$ in~$T$,
  \item $cv^{T,S}_{i,j}$ denotes the number of common leaves of the vertical subtree of $i$ in $T$ and the vertical subtree of $j$ in $S$.
  \end{itemize}  
\end{lemma}
\begin{proof}
  Let $\pi$ be any permutation in~$\mathcal{L}(P_s(R))$.
  From \cref{lem:yinminmax}, for a pair $i<j \in [n]$, there exists a yin arc~$\alpha$ whose shard polytope $\SP(\alpha)$ contributes to $\b{e}_i-\b{e}_j$ if and only if $\pi^{-1}(i)>\pi^{-1}(j)$, hence if $i\succ_s j$.
  Provided this holds, the set of yin arcs $\alpha$ such that $\SP(\alpha)$ contributes to $\b{e}_i-\b{e}_j$ are defined by pairs of contiguous intervals $I,J$ of $[n]$ with $i\in I$, $j\in J$, such that $i$ is the index of the maximum of $\pi^{-1}$ in $I$, and $j$ is the index of the maximum of $\pi^{-1}$ in $J$.
  We claim that the number of choices of the three endpoints defining such a pair of intervals $I$ and $J$ is the given product.

@@

\begin{lemma}
  \label{lem:whentosum}
  Let $R$ be a rectangulation of size~$n$, let $S$ and $T$ be its source and target trees, and let~$([n],\prec_s)$ be the associated strong order.
  Given a pair $i,j\in [n]$ such that $i<j$ and $i\prec_s j$, we have
  \[
    cv^{T,S}_{i,j} > 0 \quad \Longrightarrow \quad i \verticalPattern j.
  \]
\end{lemma}
\begin{proof}
  Consider the two rectangles $r_i$ and $r_j$ with inorder labels $i$ and $j$ in $R$.
  We first observe that the source vertex $s(r_j)$ is not contained in the interior of the lower left quadrant of origin $t(r_i)$.
  Indeed, if we suppose otherwise, then the line segment from $s(r_j)$ to $t(r_i)$ has positive slope, which implies that $j\tri i$, which in turn implies $j\prec_s i$, contradicting our hypothesis.

>>>>

% Loday and anti-Loday accents (?)
\makeatletter
\newcommand{\oset}[3][0ex]{%
  \mathrel{\mathop{#3}\limits^{
    \vbox to#1{\kern-2\ex@
    \hbox{$\scriptstyle#2$}\vss}}}}
%
\newcommand{\uset}[3][0ex]{%
  \mathrel{\mathop{#3}\limits_{
    \vbox to#1{\kern-7\ex@
    \hbox{$\scriptstyle#2$}\vss}}}}
\makeatother
%
\newcommand{\loday}[1]{\smash{\overset{\frown}{#1}}}
\newcommand{\antiloday}[1]{\smash{\overset{\smile}{#1}}}
\newcommand{\upArc}[1]{\smash{\raisebox{.05cm}{$\uset[0ex]{#1}{\frown}$}}}
\newcommand{\downArc}[1]{\smash{\raisebox{-.05cm}{$\oset[.2ex]{#1}{\smile}$}}}
% yin and yang accents (?)
\newcommand{\yin}[1]{\smash{\overset{\sim}{#1}}}
\newcommand{\yang}[1]{\smash{\overset{\backsim}{#1}}}
\newcommand{\yinArc}[2]{\smash{\raisebox{.05cm}{$\uset[0ex]{#1}{\frown}$}\hspace{-.3ex}\raisebox{-.03cm}{$\oset[.2ex]{#2}{\smile}$}}}
\newcommand{\yangArc}[2]{\smash{\raisebox{-.03cm}{$\oset[.2ex]{#1}{\smile}$}\hspace{-.3ex}\raisebox{.05cm}{$\uset[0ex]{#2}{\frown}$}}}
% weak and strong equivalence on S_n
\newcommand{\weakeq}{\asymp}
\newcommand{\strongeq}{\mathbin{\smash{\begin{smallmatrix} \backsim \\[-.3cm] \sim \end{smallmatrix}}}}

@@

Given a rectangulation~$R$, let us consider the source and target trees~$S(R)$ and~$T(R)$, and orient the horizontal edges from left to right, and the vertical edges from bottom to top.
This defines two partial orders on $[n]$.
The \defn{weak poset}~$P_w(R)=([n],\preccurlyeq_w)$ of the rectangulation~$R$ is the transitive closure of the union of these two partial orders defined by~$S(R)$ and~$T(R)$.
Equivalently, the weak poset can be defined by considering the adjacency graph of the rectangles in a diagonal representative of the weak equivalence class of $R$, orienting its edges from left to right and from bottom to top, and taking the transitive closure of this directed acyclic graph.
These posets were characterized (and called Baxter posets) by E.~Meehan~\cite{MR4014603}.
The weak poset is a two-dimensional lattice, whose minimum is the root of~$S(R)$ and whose maximum is the root of~$T(R)$.
See \cref{fig:weakPoset}.

@@

We call this congruence the \defn{weak rectangulation congruence} (it is sometimes referred to as the \defn{Baxter congruence}).
It is the intersection of the sylvester congruence with the antisylvester congruence.
The \defn{weak rectangulotope} $\WRP(n)$ is the quotientope of the weak rectangulation congruence $\weakeq$ on the weak Bruhat order on $\f{S}_n$.

By construction, the quotient fan~$\c{F}_\weakeq$ has a maximal cone~$C_R$ for each weak rectangulation~$R$ of size~$n$.
\vincent{Notations? $\preccurlyeq_R^w$ and~$C_R^w$?}
This cone is given by~$C_R = \set{\b{x} \in \R^n}{x_i \le x_j \text{ if } i \preccurlyeq_w j}$, where~$\preccurlyeq_w$ is the weak poset of~$R$.
Note that there is a bijection between the facets of~$C_R$, the cover relations of~$\preccurlyeq_w$, and the flips from~$R$.
The definition of the weak poset just translates the fact that the weak rectangulation fan~$\c{F}_\weakeq$ is the common refinement of the sylvester fan and the anti-sylvester fan.
Namely, as~$C_R$ is the intersection of the cone of the sylvester fan corresponding to the target tree~$T(R)$ with the cone of the anti-sylvester fan corresponding to the source tree~$S(R)$, the weak poset~$\preccurlyeq_w$ is the transitive closure of the union of the posets associated to~$S(R)$ and~$T(R)$.

Note that weak rectangulotopes were first constructed by S.~Law and N.~Reading in~\cite{MR2871762} as Minkowski sums of two opposite associahedra.
Our vertex and facet descriptions of \cref{thm:weakRectangulotope} can be directly derived from this construction.
Here, we provide a more pedestrian approach based on Minkowski sums of shard polytopes, as it will serve as a template for the more involved proof of \cref{thm:strongRectangulotope} in \cref{sec:strongRectangulotopes}.

@@

These arcs can be classified into two families of arcs: the up arcs and the down arcs.
For an interval $I$ of $[n]$ of size at least~$2$, the \defn{up arc}~$\upArc{I}$ (resp.~\defn{down arc}~$\downArc{I}$) is the arc that starts at~$\min(I)$ and ends at~$\max(I)$, and passes above (resp.~below) all remaining elements of~$I$, that is
\begin{align*}
\upArc{I} & \eqdef \big( \min(I), \max(I), I \ssm \{\min(I), \max(I)\}, \varnothing \big) \\
\qquad\text{and}\qquad
\downArc{I} & \eqdef \big( \min(I), \max(I), \varnothing, I \ssm \{\min(I), \max(I)\} \big).
\end{align*}
Note that the basic arc~$(i, i+1, \varnothing, \varnothing)$ is both an up and down arc.
The up arcs correspond to the sylvester congruence, while the down arcs correspond to the anti-sylvester congruence.

@@

Given a rectangulation $R$, we define the \defn{strong poset} $P_s(R)=([n],\preccurlyeq_s)$ of $R$ by considering two relations on $[n]$~\cite{ACFF24}:
\begin{enumerate}
\item The \defn{adjacency poset} $([n],\tri)$ of $R$, defined by considering the adjacency graph of the rectangles of $R$ (identified by their inorder labels in $[n]$), orienting its edges from left to right and from bottom to top, and taking the transitive closure of this directed acyclic graph.
\item The \defn{special relation} $([n],\btri)$ of $R$, in which $a\btri b$ if the two rectangles $r_a$ and $r_b$ with respective inorder labels $a$ and $b$ are such that 
	\begin{enumerate}[(i)]
	\item either the right side of $r_b$ and the left side of $r_a$ lie on the same vertical segment, and the bottom-right corner of $r_b$ is above the top-left corner of $r_a$, 
	\item or the top side of $r_b$ and the bottom side of rectangle $r_a$ lie on the same horizontal segment, and the top-left corner of $r_b$ lies on the right of the bottom-right corner~of~$r_a$.
	\end{enumerate}
\end{enumerate}
The strong poset is the transitive closure of the union of the two relations $\tri$ and $\btri$.
See \cref{fig:strongPoset}.
The strong poset is a two-dimensional lattice, whose minimum is the root of~$S(R)$ and whose maximum is the root of~$T(R)$. The following statement will be used later.

@@

By construction, the quotient fan~$\c{F}_{\strongeq}$ has a maximal cone~$C_R$ for each strong rectangulation~$R$ of size~$n$.
This cone is given by~$C_R = \set{\b{x} \in \R^n}{x_i \le x_j \text{ if } i \preccurlyeq_s j}$, where~$\preccurlyeq_s$ is the strong poset of~$R$.
Note that there is a bijection between the facets of~$C_R$, the cover relations of~$\preccurlyeq_s$, and the flips from~$R$.

@@

\cref{prop:strongMinkowski} gives us an alternative description of $\SRP(n)$ as the sum of two opposite quotientopes defined by the \defn{yin} and \defn{yang congruences} respectively, where the yin (resp.~yang) congruence is defined by the arc ideal consisting of all yin (resp.~yang) arcs.
It can be checked that the equivalence classes of these congruences are counted by \defn{semi-Baxter permutations}, defined as permutations avoiding $2\underline{41}3$, hence only one of the two forbidden Baxter patterns.
These equivalence classes are in turn in bijection with rectangulations that avoid one of the two patterns defining diagonal rectangulations, see~\cite{ACFF24}.
\vincent{Move this}

@@

\begin{lemma}
  \label{lem:yincount}
  Let $R$ be a strong rectangulation of size~$n$, let~$S$ and $T$ be its source and target trees, let~$([n],\preccurlyeq_s)$ be its strong poset, and let~$\b{p}(R)$ be the vertex of $\SRP(n)$ corresponding to $R$.
  Given a pair $i,j\in [n]$ with $i<j$, the number of yin arcs $\alpha$ such that $\b{e}_i-\b{e}_j$ is the extremal vertex of~$\SP(\alpha)$ contributing to $\b{p}(R)$ is
  \[
    h^T_i \cdot cv^{T,S}_{i,j}\cdot h^S_j 
  \]
  if $i\succ_s j$, and $0$ otherwise, where
  \begin{itemize}
  \item $h^T_i$ denotes the number of leaves in the horizontal and vertical subtrees of $i$ in~$T$,
  \item $cv^{T,S}_{i,j}$ denotes the number of common leaves of the vertical subtree of $i$ in $T$ and the vertical subtree of $j$ in $S$.
  \end{itemize}  
\end{lemma}
\begin{proof}
  Let $\pi$ be any permutation in~$\mathcal{L}(P_s(R))$.
  From \cref{lem:yinminmax}, for a pair $i<j \in [n]$, there exists a yin arc~$\alpha$ whose shard polytope $\SP(\alpha)$ contributes to $\b{e}_i-\b{e}_j$ if and only if $\pi^{-1}(i)>\pi^{-1}(j)$, hence if $i\succ_s j$.
  Provided this holds, the set of yin arcs $\alpha$ such that $\SP(\alpha)$ contributes to $\b{e}_i-\b{e}_j$ are defined by pairs of contiguous intervals $I,J$ of $[n]$ with $i\in I$, $j\in J$, such that $i$ is the index of the maximum of $\pi^{-1}$ in $I$, and $j$ is the index of the maximum of $\pi^{-1}$ in $J$.
  We claim that the number of choices of the three endpoints defining such a pair of intervals $I$ and $J$ is the given product.

@@

\begin{lemma}
  \label{lem:whentosum}
  Let $R$ be a rectangulation of size~$n$, let $S$ and $T$ be its source and target trees, and let~$([n],\preccurlyeq_s)$ be the associated strong order.
  Given a pair $i,j\in [n]$ such that $i<j$ and $i\prec_s j$, we have
  \[
    cv^{T,S}_{i,j} > 0 \quad \Longrightarrow \quad i \verticalPattern j.
  \]
\end{lemma}
\begin{proof}
  Consider the two rectangles $r_i$ and $r_j$ with inorder labels $i$ and $j$ in $R$.
  We first observe that the source vertex $s(r_j)$ is not contained in the interior of the lower left quadrant of origin $t(r_i)$.
  Indeed, if we suppose otherwise, then the line segment from $s(r_j)$ to $t(r_i)$ has positive slope, which implies that $j\tri i$, which in turn implies $j\prec_s i$, contradicting our hypothesis.
